---
title: "BTEP-DateTime"
author: "Drs. Sarangan Ravichandran and Randall Johnson"
date: "February 26, 2017" 
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "H:/2017/BTEP1-TidyingData")
```
Let us load the libraries
```{r loadlibraries}
library(tidyverse) 
```


## Date & Times in R

Dates and Times are very important. 
Routinely used in expression (gene) data


```{r}
D1 <- as.Date('1920-6-16')
class(D1)

D2 <- as.Date('2017/03/07')
class(D2)

## Current Date from System 
Sys.Date()

```

If you use <cod> ?as.date </code>, you will see the formats that 
the function takes. 
```{r kable}
Code <- c("%d", "%m","%b","%B","%y","%Y")
Value <- c("Day of the month (decimal numeber)", "Month(decimal number)", "Month (abbreviation)",
           "Month(full name)", "Year (2 digit)", "Year (4 digit)")
library(knitr)
tibble1 <- tibble(Code, Value)
kable(tibble1)

```

The function <pre> <code> as.Date </code> </pre> expects the date to be in the - separated format. 
If it is not, then we should use the following format

```{r}
as.Date('03/07/2017', format='%m/%d/%Y')
as.Date('March 07, 2017', format='%B %d, %Y')
```

Let us look to see how R stores the data internally.
Two different datatypes that deal with Time and Date variables: 

## POSIXct

Will provide you the number of seconds that has been elapsed since January 1, 1970. 

Negative numebrs indicate seconds before this time. 

## POSIXlt

- Type of vector

```{r kable1}
Position <- 1:9
Value <- c("seconds","minutes","hours","day of the month (1-31)", "month of the year (0-11)", 
           "years since 1900", "day of the week (0: Sunday - 6: Saturday)", "Day of the year (0-365)",
           "Daylight savings indicator (Positive/Negative: Yes/No")
library(knitr)
tibble2 <- tibble(Position, Value)
kable(tibble2)

```

Let us see how we can use these two data types:

```{r}
(timext <- Sys.time())
class(timext)
typeof(timext)
cat(timext, "\n")

(timelt <- as.POSIXlt(timext) )
typeof(timelt)
names(timelt)
lapply(timelt, function(x) print(x))
```

 ## How can we use the time? 
 
```{r}
 timenow <- Sys.time()
timenow
timenow - 30 # 30 seconds earlier 

```
Be careful with time data that are used for numerical calculations. 


# How to store Time in different formats and use it at a later time or using a differnt software?

Two commands that can help us accomplish this, strftime and strptime.

strftime: converts a time data and convert it to a string
strptime: converts a sting and convert it to Date/time format suitable for R calculations
```{r}
string_timenow <- strftime(timenow, "%Y-%m-%d %H:%M:%S")  #"2017-03-07 15:52:03 EST"
string_timenow
class(string_timenow)
class(timenow)
typeof(string_timenow)
(Posixct_timenow <- strptime(string_timenow,"%Y-%m-%d %H:%M:%S" ))
class(Posixct_timenow)
```


### <span style="color:green">Final R-basic exercise</span>

The datset that we will be using for this section comes from the  
National Electronic Injury Surveillance System (NEISS)
https://www.cpsc.gov/research--statistics/neiss-injury-data 
Here is a short description of the data file from the NEISS. 

*"Each record (case) is separated by a carriage return/line feed, and the
fields (parameters and narrative) are separated by a tab character, which you can specify as the delimiter when importing into a spreadsheet or database."*

Before you read in the data, please take a look at the following two files: **NEISS_SAS_formats.txt** and **NEISS_SAS_variance.txt** 
These two files will act as the codebook for the data. 

Read the data file, **nss15.tsv** file from the sub-folder/directory, **Data** and call the data as **nss15** variable.

**Hint:** Use the File --> Import --> Data DataSet option to read the file. 
To remind the data is in a tab-separated format.

**Hint:** Watch out for the Data Type choices that are suggested to you and choose the appropriate ones

**Hint:** Please watch out for any warnings or issues while R/Rstudio 
is reading the files. If you spot any errors, think of how to fix it.
After fixing the problems (if any), go back and read the file. 
 
Answer the following questions.

- Report the number of cases for the month of May? 
  Tell us how many cases were reported for May 13 - May 16, 2015? 
  Use this information to answer the following questions. 
- What hospitals were the cases went to (provide a table)?
- How many were children ( < 5 years)? 
- provide the proportion of male/female? 
- What was the race distribution? 


