---
title: "Prework"
date: "2/28/2018"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prerequisites

We are assuming that you have some familiarity with R before attending this hands-on seminar. Please review our [R/RStudio Intro](https://github.com/ravichas/TidyingData/blob/master/0-RStudio-Intro.md) before coming, and if you are unfamiliar with the material covered there or if you would like to brush up on your basic R skills before attending, you may find the [DataCamp](https://www.datacamp.com/courses/free-introduction-to-r) or [TryR](http://tryr.codeschool.com) tutorials helpful. Both of these tutorials pretty much cover the same material.

There is also a nice tutorial within the `swirl` package. To start this up, run the following commands within RStudio:

```{r eval=FALSE}
install.packages("swirl")
library(swirl)
swirl()
```

## Computer setup

Computers in the training room will have R, RStudio, and the packages we'll be using installed, but if you would like to bring your own computer, please install the following:

* [R](https://cran.r-project.org)
* [RStudio](https://www.rstudio.com/products/rstudio/download/)
* R packages (you can install these using the code below):
    * [tidyverse](https://www.tidyverse.org/)
    * [biobroom](https://bioconductor.org/packages/release/bioc/html/biobroom.html)

To check for the required packages and to install any missing packages, you can run this code within the RStudio console window:

```{r message = FALSE}
if(!require(tidyverse))
    install.packages('tidyverse')

if(!require(biobroom))
{
    # if this fails, sourcing http://bioconductor.org/biocLite.R might work
    source("https://bioconductor.org/biocLite.R")
    biocLite("biobroom")
}
```

To verify that you have correctly set up your system, run this line of code in RStudio. It will either print a message to your screen indicating your computer is setup properly, or it will throw an error with information on missing packages.

```{r check sysem}
source('https://raw.githubusercontent.com/ravichas/TidyingData/master/Prework/systemCheck.R')
```
